Modeling the impacts of park access on health outcomes: a utility-based accessibility approach

### 2020-06-05 manuscript/submission/manuscript.tex
De-identify manuscript and write modified text in blue
Also include response to reviewers document

\author{Gregory Macfarlane\affilnum{1}, Nico Boyd\affilnum{2}, John E. Taylor\affilnum{3}, Kari Watkins\affilnum{3}}

\affiliation{\affilnum{1}{Department of Civil \& Environmental Engineering, Brigham Young University}\\\affilnum{2}{Fehr \& Peers, Los Angeles California}\\\affilnum{3}{School of Civil \& Environmental Engineering, Georgia Institute of Technology}}

\corrauth{Gregory Macfarlane, 430 Engineering Building, Provo UT 84663}

\email{\href{mailto:gregmacfarlane@byu.edu}{\nolinkurl{gregmacfarlane@byu.edu}}}

@@

Historically, most large-scale health promotion efforts focused on
individual-level interventions intended to educate people about healthy
lifestyles and behaviors, touching on topics including diet and
exercise. Over the last several years a new \emph{social} model of
health has evolved, described by \citet{Duhl1999} ``as an outcome of the
effects of socioeconomic status, culture, environmental conditions,
housing, employment and community
influences''\citeyearpar[p.7]{Duhl1999}. In this new paradigm, resources
provided via civil infrastructure --- in particular parks and other
public green spaces --- play a crucial role in promoting and sustaining
public health \citep{Bedimo-Rung2005, Wells2007, Coutts2008}. After all,
it does little to educate individuals on the importance of exercise if
their community is designed and built in such a way that such exercise
is expensive, unenjoyable, or unsafe. It is similarly necessary for
urban planners to identify which neighborhoods have access to quality
green space, and which neighborhoods need intervention in the form of
new or rehabilitated green space, or improved transportation access to
existing quality green space.

@@

\label{sec:litreview} The question of how green space in the urban
environment affects the physical, emotional, and mental well-being of
urban residents is not new to the academic literature. Comprehensive
reviews and analysis of this literature up to 2007 can be had from
\citet{tzoulas2007promoting}, and for the subsequent decade from
\citet{kabisch2015human}. In spite of the depth of this research arena
--- or perhaps because of it --- quantitative approaches for assessing
both health outcomes and access to green space vary widely. It is
therefore not surprising that in a meta-analysis of 20 recent peer
reviewed journal articles exploring the relationship between access to
parks and objectively measured physical activity, \citet{Bancroft2015}
found that five studies reported a significant positive association, six
studies produced mixed results, and nine studies found no association at
all.

Loosely defined, the \emph{accessibility} of an arbitrary place
describes the ease with which people can accomplish activities there.
Accessibility is an abstract concept with tempting quantifiability
\citep{Handy1997}, but objective quantification may not be strictly
necessary. For example, \citet{takano2002urban} showed improved survival
rates among senior citizens in Japan who self-reported having access to
qualitatively good walking spaces. Though encouraging, this result could
simply reveal self-selection: Those seniors who felt the walking space
was good were those who chose to enjoy it. Urban planners seeking to
evaluate which residents have good access to quality green space may
desire objective and therefore comparable measures.

@@

Applications of this isochrone-based framework are easily derived and
relatively common. ParkScore \citep{parkscore2019} is a sophisticated
application of this approach where \(D\) is a carefully calculated
10-minute network-bound walk, paired with a demographic analysis of
areas within and outside of this threshold. \citet{devries2003natural}
showed the amount of greenspace within a 3-kilometer radius was
positively associated with a suite of self-reported health outcomes.
Similarly, \citet{devries2013streetscape} showed that streetscape
greenery within 500 meters of a residence was associated with perceived
general health. Conversely, \citet{carlson2012interactions} found no
relationship between the health of senior citizens and the number of
parks within 500 meters.

Another common access measure fits in this category, where
distance-based isochrones are replaced with political, geographical, or
statistical boundaries. For example, \citet{Mitchell2008} find that
inequalities in health outcomes attributable to income are mitigated in
statistical areas with a greater percentage of green space. These are
especially common in comparative macroscopic studies, such as the
positive relationships found by \citet{West2012} between density of park
land in a metropolitan area and self-reported population health
indicators derived from the Behavioral Risk Factor Surveillance System
(BRFSS). On the other hand, \citet{Larson2016} found no relationship
between physical health and park quantity, quality, and accessibility in
44 U.S. cities and self-reported scores on the Gallup-Healthways
Wellbeing Index. At the sub-metropolitan level, \citet{Stark2014} found
a significant relationship in New York City between the percent of a
residential ZIP code that is park space and objectively measured body
mass index (BMI). It is also possible to attenuate an isochrone with
various factors; for instance, \citet{Dias2019} considered road safety
perception as an extenuating factor in the effective distance to a park.

In spite of the flexibility of adapted isochrone techniques, the
arbitrary definitions necessarily imposed by researchers in its
application can inhibit holistic analysis \citep{Logan2017}. A somewhat
more complete approach is the so-called ``gravity'' accessibility
statistic. In this case the accessibility of point \(i\) is the
denominator of the gravity formulation of a trip distribution model,
\begin{equation}\label{eq:gravity}
A_i = \sum_{j = 1}^J S_j f(d_{ij})
\end{equation} where \(S_j\) is the ``size'' of each park literally
(i.e.~acres) or abstractly (i.e. trip attractions) and \(f(d_{ij})\) is
a monotonically decreasing cost function, usually a negative
exponential. \citet{Dong2006} note that the isochrone framework is a
special case of the gravity model where \(f(d_{ij})\) is a binary
function and \(S_j = 1\). \citet{Giles-Corti2005} compare various
gravity-based accessibility scores with an isochrone specification and
show that the former are more predictive of walking behavior; that is,
the attributes of a park influence walking more than merely the distance
to it. \citet{hillsdon2006relationship} found no relationship between a
gravity-based accessibility term and self-reported physical activity
among middle-aged individuals in Norwich, England. \citet{Zhang2011}
developed a population-weighted, gravity-based accessibility to parks
metric with a national scope, though they do not examine its correlation
with health outcomes.

@@

We predict either the physical activity or the obesity rate (\(y\)) in a
census tract as a linear function of the tract's sociodemographic
characteristics \(X\) and accessibility to parks \(\boldsymbol{A}\),
\begin{equation}\label{eq:themodel}
 \boldsymbol{y} = X\boldsymbol{\beta} + \beta_{a}\boldsymbol{A} + \boldsymbol{\varepsilon}
\end{equation} where \(X\) is a matrix of dimension \(n\times p\) where
\(n\) is the number of observations and \(p\) is the number of predictor
variables, and \(\boldsymbol{\beta}\) is a \(p\times 1\) vector of
estimated parameters. The accessibility vector \(\boldsymbol{A}\)
influences physical activity or obesity rates by the single estimated
parameter \(\beta_a\), and a vector \(\boldsymbol{\varepsilon}\)
represents the model residual as a random variable.

@@

To determine which spatial effects specification was appropriate, we
estimated SLX and SDEM models of physical activity participation rates
as a function of tract covariates and accessibility. These two models
are presented in Table \ref{tab:slx-sdem}. With only a few exceptions,
the SLX and SDEM coefficient estimates for corresponding covariates lie
within the 95\% confidence intervals of the other model. A Hausman-style
test \citep{Pace2008} on the coefficients assumes a null hypothesis that
the coefficients are equivalent; this test produces \(p\)-value of only
\(1.307e-23\). Though this statistic is well beneath a common hypothesis
test threshold of \(\alpha = 0.05\), we elect retain the null hypothesis
that the coefficients are equivalent, as the differences are at no point
practically different from each other. The consequences of rejecting the
null hypothesis in this case would be to suggest a model with spatial
dependence of the dependent variable; as stated above, we feel that such
a relationship is unreasonable. As a consequence of these analyses, we
adopt the SDEM specification moving forward as the spatial lag of the
errors is significant and the overal log likelihood of the SDEM is
higher than the SLX specification.

@@

We readily acknowledge limitations in this study. As in any study
conducted with areal data, we are at risk of falling victim to the
ecological inference fallacy, where aggregate statistics mask or
contradict disaggregated or individual-level trends. The socioeconomic
and public health data used in this study are only available at the
tract level; projecting the tract-level data to the block group level
and computing more spatially precise accessibility logsums proved
infeasible as the spatial lags of the independent variables introduced
substantial collinearity in the model equations. A large-sample survey
of individuals in New York City, including measured physical activities
and obesity would always be preferable to the tract-level data used in
this study. An ideal survey to address the question would incorporate
both sets of questions: physical activity and health data on one hand
and park use (including which parks were used and how frequently) on the
other. As no such dataset exists to our knowledge, this tract-level
aggregate analysis with asserted and transferred utility coefficients is
the possibility that remains.

@@

An analyst could then compute the utility-based accessibility for each
cluster with different utility values for each park's amenities, and
obtain a measure of a neighborhood's accessibility to park features that
its residents most care about. In this paper, we proceed only
incrementally towards this by adding some selected park amenities as an
element of each park's attractiveness above and beyond its size. The set
of amenities we included in the utility calculation --- and the
contribution of each amenity to the overall utility --- was dictated by
the availability of data on the park amenities in New York City and our
ability to find comparable prior studies with transferrable utility
parameters. We can envision a study where mobile device data revealing
the home locations of park users would be paired with more detailed park
amenities and use data (e.g.~geolocated Twitter data \emph{a la}
\citet{roberts2017using}) to estimate the utility contribution of these
features. This could in turn provide a more nuanced understanding of the
relationship between park accessibility, use, and health outcomes.

As a travel impedance measure, we used the Euclidean distance between
each tract's population-weighted centroid and the nearest point on the
edge of a park's border. Euclidean distances have well-rehearsed
limitations regarding their fidelity with the underlying infrastructure
network, etc. Network-based distances can also suffer from challenges
when applied to multimodal problems; these challenges are exacerbated
when the non-highway mode share is high, as is likely when considering
access to parks. We examined replacing the Euclidean distance with a
network-based distance; the choice of distance metric appeared
inconsequential to our larger findings in this case while substantially
increasing the computational load required. Additionally, we could not
verify the accuracy of the underlying network data at the scale of this
analysis. We therefore retain the Euclidean distance for simplicity. A
better metric of travel impedance may be a mode choice model logsum,
which weights all alternative travel alternatives against each other. We
leave this as a recommendation for future research.

@@

Finally, this study is primarily focused on the hypothesis that
accessibility to parks encourages physical activity, which in turn
reduces obesity. There are a multitude of other hypotheses that might be
proposed and tested with the basic methodology we have employed here, or
competing explanations for the outcomes we have observed. It is
distinctly possible, for instance, that individuals who wish to exercise
regularly in parks choose to live near them. Given that the CDC models
informing the small-area obesity and physical activity estimates
presumably include variables likely to influence such preferences
(income, etc.), our investigation cannot isolate the preferences from
the effect. Controlling for such a self-selection effect would be
necessary to isolate the exogenous impacts of park access on obesity or
other health outcomes. And regarding these other variables: this study
did not consider potential relationships between park access and
hospitalization rates, life expectancy, respiratory disease, mental
health, or any number of potential beneficial outcomes hypothesized or
explored in the existing literature. The COVID-19 pandemic and its
associated social distancing behaviors in many regions of the world has
also drawn attention to problems of park access, use, and equity.
Exploring these connections and their underlying mechanisms should be a
priority as city planners and urban architects attempt to improve the
quality of life of urban residents in the future.

@@

\begin{acks}
This project was funded by the Speedwell Foundation. We are grateful to Rachel
Samuels and Neda Madi for helping to assemble the data.
\end{acks}

>>>>

\author{XXXX\affilnum{1}, XXXX\affilnum{2}, XXXX\affilnum{3}, XXXXX \affilnum{3}}

\affiliation{\affilnum{1}{XXXX}\\\affilnum{2}{XXXX}\\\affilnum{3}{XXXX}}

\corrauth{XXXXX}

\email{XXXX}

@@

Historically, most large-scale health promotion efforts focused on
individual-level interventions intended to educate people about healthy
lifestyles and behaviors, touching on topics including diet and
exercise. Over the last several years a new \emph{social} model of
health has evolved, described by \citet{Duhl1999} ``as an outcome of the
effects of socioeconomic status, culture, environmental conditions,
housing, employment and community
influences''\citeyearpar[p.7]{Duhl1999}. In this new paradigm, resources
provided via civil infrastructure --- in particular parks and other
public green spaces --- play a crucial role in promoting and sustaining
public health \citep{Bedimo-Rung2005, Wells2007, Coutts2008}. After all,
it does little to educate individuals on the importance of exercise if
their community is designed and built in such a way that such exercise
is expensive, unenjoyable, or unsafe. \textcolor{blue}{It is similarly necessary for
urban planners to identify which neighborhoods have access to quality
green space, and which neighborhoods need intervention in the form of
new or rehabilitated green space, or improved transportation access to
existing quality green space.}

@@

\label{sec:litreview}

\textcolor{blue}{
The question of how green space in the urban
environment affects the physical, emotional, and mental well-being of
urban residents is not new to the academic literature. Comprehensive
reviews and analysis of this literature up to 2007 can be had from
\citet{tzoulas2007promoting}, and for the subsequent decade from
\citet{kabisch2015human}. In spite of the depth of this research arena
--- or perhaps because of it --- quantitative approaches for assessing
both health outcomes and access to green space vary widely. It is
therefore not surprising that in a meta-analysis of 20 recent peer
reviewed journal articles exploring the relationship between access to
parks and objectively measured physical activity, \citet{Bancroft2015}
found that five studies reported a significant positive association, six
studies produced mixed results, and nine studies found no association at
all.}

\textcolor{blue}{
Loosely defined, the \emph{accessibility} of an arbitrary place
describes the ease with which people can accomplish activities there.
Accessibility is an abstract concept with tempting quantifiability
\citep{Handy1997}, but objective quantification may not be strictly
necessary. For example, \citet{takano2002urban} showed improved survival
rates among senior citizens in Japan who self-reported having access to
qualitatively good walking spaces. Though encouraging, this result could
simply reveal self-selection: Those seniors who felt the walking space
was good were those who chose to enjoy it. Urban planners seeking to
evaluate which residents have good access to quality green space may
desire objective and therefore comparable measures.}

@@

Applications of this isochrone-based framework are easily derived and
relatively common. ParkScore \citep{parkscore2019} is a sophisticated
application of this approach where \(D\) is a carefully calculated
10-minute network-bound walk, paired with a demographic analysis of
areas within and outside of this threshold. \textcolor{blue}{\citet{devries2003natural}
showed the amount of greenspace within a 3-kilometer radius was
positively associated with a suite of self-reported health outcomes.
Similarly, \citet{devries2013streetscape} showed that streetscape
greenery within 500 meters of a residence was associated with perceived
general health. Conversely, \citet{carlson2012interactions} found no
relationship between the health of senior citizens and the number of
parks within 500 meters.}

Another common access measure fits in this category, where
distance-based isochrones are replaced with political, geographical, or
statistical boundaries. \textcolor{blue}{For example, \citet{Mitchell2008} find that inequalities
in health outcomes attributable to income are mitigated in statistical areas
with a greater percentage of green space. \citet{Stark2014} found a significant
relationship in New York City between the percent of a residential ZIP code that
is park space and objectively measured body mass index (BMI).} Political or
regional boundaries are
especially common in comparative macroscopic studies, such as the
positive relationships found by \citet{West2012} between density of park
land in a metropolitan area and self-reported population health
indicators derived from the Behavioral Risk Factor Surveillance System
(BRFSS). On the other hand, \citet{Larson2016} found no relationship
between physical health and park quantity, quality, and accessibility in
44 U.S. cities and self-reported scores on the Gallup-Healthways
Wellbeing Index. It is also possible to attenuate an isochrone with
various factors; for instance, \citet{Dias2019} considered road safety
perception as an extenuating factor in the effective distance to a park.

In spite of the flexibility of adapted isochrone techniques, the
arbitrary definitions necessarily imposed by researchers in its
application can inhibit holistic analysis \citep{Logan2017}. A somewhat
more complete approach is the so-called ``gravity'' accessibility
statistic. In this case the accessibility of point \(i\) is the
denominator of the gravity formulation of a trip distribution model,
\begin{equation}\label{eq:gravity}
A_i = \sum_{j = 1}^J S_j f(d_{ij})
\end{equation} where \(S_j\) is the ``size'' of each park literally
(i.e.~acres) or abstractly (i.e. trip attractions) and \(f(d_{ij})\) is
a monotonically decreasing cost function, usually a negative
exponential. \citet{Dong2006} note that the isochrone framework is a
special case of the gravity model where \(f(d_{ij})\) is a binary
function and \(S_j = 1\). \citet{Giles-Corti2005} compare various
gravity-based accessibility scores with an isochrone specification and
show that the former are more predictive of walking behavior; that is,
the attributes of a park influence walking more than merely the distance
to it. \textcolor{blue}{\citet{hillsdon2006relationship} found no relationship between a
gravity-based accessibility term and self-reported physical activity
among middle-aged individuals in Norwich, England.} \citet{Zhang2011}
developed a population-weighted, gravity-based accessibility to parks
metric with a national scope, though they do not examine its correlation
with health outcomes.

@@

We predict either the physical activity or the obesity rate (\(y\)) in a
census tract as a linear function of the tract's sociodemographic
characteristics \(X\) and accessibility to parks \(\boldsymbol{A}\),
\begin{equation}\label{eq:themodel}
 \boldsymbol{y} = X\boldsymbol{\beta} + \beta_{a}\boldsymbol{A} + \boldsymbol{\varepsilon}
\end{equation}
\textcolor{blue}{where \(X\) is a matrix of dimension \(n\times p\) where
\(n\) is the number of observations and \(p\) is the number of predictor
variables, and \(\boldsymbol{\beta}\) is a \(p\times 1\) vector of
estimated parameters. The accessibility vector \(\boldsymbol{A}\)
influences physical activity or obesity rates by the single estimated
parameter \(\beta_a\), and a vector \(\boldsymbol{\varepsilon}\)
represents the model residual as a random variable.}

@@

To determine which spatial effects specification was appropriate, we
estimated SLX and SDEM models of physical activity participation rates
as a function of tract covariates and accessibility. These two models
are presented in Table \ref{tab:slx-sdem}. With only a few exceptions,
the SLX and SDEM coefficient estimates for corresponding covariates lie
within the 95\% confidence intervals of the other model. A Hausman-style
test \citep{Pace2008} on the coefficients assumes a null hypothesis that
the coefficients are equivalent; this test produces \(p\)-value of only
\(1.307e-23\). Though this statistic is well beneath a common hypothesis
test threshold of \(\alpha = 0.05\), we elect to retain the null hypothesis
that the coefficients are equivalent, as the differences are at no point
practically different from each other. The consequences of rejecting the
null hypothesis in this case would be to suggest a model with spatial
dependence of the dependent variable; as stated above, we feel that such
a relationship is unreasonable. As a consequence of these analyses, we
adopt the SDEM specification moving forward as the spatial lag of the
errors is significant and the overal log likelihood of the SDEM is
higher than the SLX specification.

@@

We readily acknowledge limitations in this study. As in any study
conducted with areal data, we are at risk of falling victim to the
ecological inference fallacy, where aggregate statistics mask or
contradict disaggregated or individual-level trends. \textcolor{blue}{The socioeconomic
and public health data used in this study are only available at the
tract level; projecting the tract-level data to the block group level
and computing more spatially precise accessibility logsums proved
infeasible as the spatial lags of the independent variables introduced
substantial collinearity in the model equations.} A large-sample survey
of individuals in New York City, including measured physical activities
and obesity would always be preferable to the tract-level data used in
this study. An ideal survey to address the question would incorporate
both sets of questions: physical activity and health data on one hand
and park use (including which parks were used and how frequently) on the
other. As no such dataset exists to our knowledge, this tract-level
aggregate analysis with asserted and transferred utility coefficients is
the possibility that remains.

@@

An analyst could then compute the utility-based accessibility for each
cluster with different utility values for each park's amenities, and
obtain a measure of a neighborhood's accessibility to park features that
its residents most care about. In this paper, we proceed only
incrementally towards this by adding some selected park amenities as an
element of each park's attractiveness above and beyond its size. \textcolor{blue}{The set
of amenities we included in the utility calculation --- and the
contribution of each amenity to the overall utility --- was dictated by
the availability of data on the park amenities in New York City and our
ability to find comparable prior studies with transferrable utility
parameters. We can envision a study where mobile device data revealing
the home locations of park users would be paired with more detailed park
amenities and use data (e.g.~geolocated Twitter data \emph{a la}
\citet{roberts2017using}) to estimate the utility contribution of these
features.} This could in turn provide a more nuanced understanding of the
relationship between park accessibility, \textcolor{blue}{use,} and health outcomes.

As a travel impedance measure, we used the Euclidean distance between
each tract's population-weighted centroid and the nearest point on the
edge of a park's border. Euclidean distances have well-rehearsed
limitations regarding their fidelity with the underlying infrastructure
network, etc. Network-based distances can also suffer from challenges
when applied to multimodal problems; these challenges are exacerbated
when the non-highway mode share is high, as is likely when considering
access to parks. \textcolor{blue}{We examined replacing the Euclidean distance with a
network-based distance; the choice of distance metric appeared
inconsequential to our larger findings in this case while substantially
increasing the computational load required. Additionally, we could not
verify the accuracy of the underlying network data at the scale of this
analysis. We therefore retain the Euclidean distance for simplicity.} A
better metric of travel impedance may be a mode choice model logsum,
which weights all alternative travel alternatives against each other. We
leave this as a recommendation for future research.

@@

Finally, this study is primarily focused on the hypothesis that
accessibility to parks encourages physical activity, which in turn
reduces obesity. There are a multitude of other hypotheses that might be
proposed and tested with the basic methodology we have employed here, or
competing explanations for the outcomes we have observed. It is
distinctly possible, for instance, that individuals who wish to exercise
regularly in parks choose to live near them. Given that the CDC models
informing the small-area obesity and physical activity estimates
presumably include variables likely to influence such preferences
(income, etc.), our investigation cannot isolate the preferences from
the effect. Controlling for such a self-selection effect would be
necessary to isolate the exogenous impacts of park access on obesity or
other health outcomes. And regarding these other variables: this study
did not consider potential relationships between park access and
hospitalization rates, life expectancy, respiratory disease, mental
health, or any number of potential beneficial outcomes hypothesized or
explored in the existing literature. \textcolor{blue}{The COVID-19 pandemic and its
associated social distancing behaviors in many regions of the world has
also drawn attention to problems of park access, use, and equity.}
Exploring these connections and their underlying mechanisms should be a
priority as city planners and urban architects attempt to improve the
quality of life of urban residents in the future.

@@

\begin{acks}
\end{acks}
